---
title: "Introduction to `lingtypology`"
author: "[G. Moroz](mailto:agricolamz@gmail.com) <br> Presentation link: [tinyurl.com/y9lbrzf6](https://tinyurl.com/y9lbrzf6)"
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE, comment=NA, fig.width=8, fig.height=3.4)
library(lingtypology)
```

## 1. What is `lingtypology`?

* simple tool for creating interactive and static maps in R
* you can use the result on your website and in your projects
* API to some linguistic databases (Glottolog, WALS, PHOIBLe, AUTOTYP and others)

## 2. Package instalation

Get the stable version from CRAN:

```{r, eval = FALSE}
install.packages("lingtypology")
```

… or get the development version from GitHub:

```{r, eval = FALSE}
# install.packages("devtools") # 
devtools::install_github("ropensci/lingtypology")
```

If you've got some problems with dependencies, try to install one of the older vertions:

```{r, eval = FALSE}
devtools::install_version("lingtypology", version = "1.0.12")
```

Misha Voronov claim that for correct instalation on Debian some additional packages should be installed:

```
apt-get install libcurl4-openssl-dev
apt-get install libssl-dev
```

Load a library:

```{r}
library(lingtypology)
```

## 2. Glottolog functions

This package is based on the [Glottolog database](http://glottolog.org/) (v. 2.7), so lingtypology has several functions for accessing data from that database.

### 2.1 Command name's syntax
Most of the functions in `lingtypology` have the same syntax: **what you need.what you have**. Most of them are based on _language name_.

* **aff.lang()** — get affiliation by language
* **area.lang()** — get macro area by language
* **country.lang()** — get country by language
* **iso.lang()** — get ISO 639-3 code by language
* **gltc.lang()** — get glottocode (identifier for a language in the Glottolog database) code by language
* **lat.lang()** — get latitude by language
* **long.lang()** — get longitude by language

Some of them help to define a vector of languages.

* **lang.aff()** — get language by affiliation
* **lang.iso()** — get language by ISO 639-3 code
* **lang.gltc()** — get language by glottocode

Additionally there are some functions to convert glottocodes to ISO 639-3 codes and vice versa:

* **gltc.iso()** — get glottocode by ISO 639-3 code
* **iso.gltc()** — get ISO 639-3 code by glottocode

[Glottolog database (v. 2.7)](http://glottolog.org/) provides `lingtypology` with language names, ISO codes, genealogical affiliation, macro area, countries, coordinates, and much information. This set of functions doesn't have a goal to cover all possible combinations of functions. Check out additional information that is preserved in the version of the Glottolog database used in `lingtypology`:

```{r}
names(glottolog.original)
```

Using R functions for data manipulation you can create your own database for your purpose.

### 2.2 Using base functions
All functions introduced in the previous section are regular functions, so they can take the following objects as input:

* a regular string
```{r}
iso.lang("Adyghe")
lang.iso("ady")
country.lang("Adyghe")
lang.aff("West Caucasian")
```

* a vector of strings
```{r}
area.lang(c("Adyghe", "Aduge"))
lang <- c("Adyghe", "Russian")
aff.lang(lang)
```

*  other functions. For example, let's try to get a vector of ISO codes for the Circassian languages
```{r}
iso.lang(lang.aff("Circassian"))
```

The behavior of most functions is rather predictable, but the function `country.lang`  has an additional feature. By default this function takes a vector of languages and returns a vector of countries.  But if you set the argument `intersection = TRUE`,  then the function returns a vector of countries where all languages from the query are spoken.
```{r}
country.lang(c("Udi", "Laz"))
country.lang(c("Udi", "Laz"), intersection = TRUE)
```

### 2.3 Spell Checker: look carefully at warnings!
There are some functions that take country names as input. Unfortunately, some countries have alternative names. In order to save users the trouble of having to figure out the exact name stored in the database (for example  _Ivory Coast_ or _Cote d'Ivoire_), all official country names and standard abbreviations are stored in the database:

```{r}
lang.country("Cape Verde")
lang.country("Cabo Verde")
head(lang.country("USA"))
```

All functions which take a vector of languages are enriched with a kind of a spell checker. If a language from a query is absent in the database, functions return a warning message containing a set of candidates with the minimal Levenshtein distance to the language from the query.

```{r}
aff.lang("Adyge")
```

### 2.4 Changes in the glottolog database
Unfortunately, the [Glottolog database (v. 2.7)](http://glottolog.org/) is not perfect for all my tasks, so I changed it a little bit. After [Robert Forkel's issue](https://github.com/ropensci/lingtypology/issues/1) I decided to add an argument `glottolog.source`, so that everybody has access to "original" and "modified" (by default) glottolog versions:

```{r}
is.glottolog(c("Abkhaz", "Abkhazian"), glottolog.source = "original")
is.glottolog(c("Abkhaz", "Abkhazian"), glottolog.source = "modified")
```

### Task 2.5: Celtic languages
How many Celtic languages in the database?

```{r, include=FALSE}
lang.aff("Celtic")
```

<form name="FormOne" onsubmit="return validateFormOne()" method="post">
<input type="text" name="answerOne">
<input type="submit" value="check">
</form>

### Task 2.6: Austronesian languages
How many Austronesian languages in the database?
```{r, include=FALSE}
length(lang.aff("Austronesian"))
```

<form name="FormTwo" onsubmit="return validateFormTwo()" method="post">
<input type="text" name="answerTwo">
<input type="submit" value="check">
</form>

### Task 2.7: Russian and Standard Arabic
What is the country where, according the database, Russian and Standard Arabic are spoken?
```{r, include=FALSE}
country.lang(c("Russian", "Standard Arabic"), intersection = TRUE)
```

<form name="FormThree" onsubmit="return validateFormThree()" method="post">
<input type="text" name="answerThree">
<input type="submit" value="check">
</form>

<script>
function validateFormOne() {
    var x = document.forms["FormOne"]["answerOne"].value;
    if (x != "6") {
        alert("I have a different answer...");
        return false;
    } else {
        alert("Exactly!");
        return false;
    }
}
function validateFormTwo() {
    var y = document.forms["FormTwo"]["answerTwo"].value;
    if (y != "1247") {
        alert("I have a different answer...");
        return false;
    } else {
        alert("Exactly!");
        return false;
    }
}
function validateFormThree() {
    var z = document.forms["FormThree"]["answerThree"].value;
    if (z != "Israel") {
        alert("I have a different answer...");
        return false;
    } else {
        alert("Exactly!");
        return false;
    }
}
</script>

## 3. Map creation

### 3.1 Base map
The most important part of the `lingtypology` package is the function `map.feature`:
```{r}
map.feature(c("Adyghe", "Kabardian", "Polish", "Russian", "Bulgarian"))
```

### 3.2 Set features
The goal of this package is to allow linguists to map language features. A list of languages and correspondent features can be stored in a `data.frame` as follows:

```{r}
df <- data.frame(language = c("Adyghe", "Kabardian", "Polish", "Russian", "Bulgarian"),
                 features = c("polysynthetic", "polysynthetic", "fusional", "fusional", "fusional"))
df
```

Now we can draw a map:
```{r}
map.feature(languages = df$language,
            features = df$features)
```

There are several types of variables in R and `map.feature` works differently depending on the variable type. I will use a build in data set `ejective_and_n_consonants` that contains 19 languages from [UPSyD database](http://www.lapsyd.ddl.ish-lyon.cnrs.fr/lapsyd/). This dataset have three variables: the categorical variable `ejectives` indicates whether some language has any ejective sound, numeric variables `consonants` and `vowels` that contains information about the number of consonants and vowels (based on UPSyD database). We can create two maps with categorical variable and with numeric variable:

```{r}
map.feature(languages = ejective_and_n_consonants$language,
            features = ejective_and_n_consonants$ejectives) # categorical
map.feature(languages = ejective_and_n_consonants$language,
            features = ejective_and_n_consonants$consonants) # numeric
```

There are two possible ways to show the World map: with the Atlantic sea or with the Pacific sea in the middle. If you don't need default Pacific view use the `map.orientation` parameter (thanks @languageSpaceLabs and @tzakharko for that idea):

```{r}
map.feature(languages = ejective_and_n_consonants$language,
            features = ejective_and_n_consonants$consonants,
            map.orientation = "Atlantic")
```

### 3.3 Set labels
An alternative way to add some short text to a map is to use the `label` option.
```{r}
map.feature(languages = df$language,
            features = df$features,
            label = df$language)
```

There are some additional arguments for customization: `label.fsize` for setting font size, `label.position` for controlling the label position, and `label.hide` to control the appearance of the label: if `TRUE`, the labels are displayed on mouse over(as on the previous map), if `FALSE`, the labels are always displayed (as on the next map).

```{r}
map.feature(languages = df$language, 
            features = df$features,
            label = df$language,
            label.fsize = 20,
            label.position = "left",
            label.hide = FALSE)
```

### Task 3.4
Create a map with Chukchi, French, Khana and Nii and add labels that don't disappear:
```{r, echo = FALSE}
map.feature(languages = c("Chukchi", "French", "Khana", "Nii"),
            label = c("Chukchi", "French", "Khana", "Nii"),
            label.hide = FALSE)
```

### Task 3.5
Create a map of Bodish languages and add minimap with argument `minimap = TRUE`:
```{r, echo = FALSE}
map.feature(languages = lang.aff("Bodish"),
            minimap = TRUE)
```