---
title: "Advanced lingtypology"
author: "[G. Moroz](mailto:agricolamz@gmail.com) <br> Presentation link: [tinyurl.com/y7vfsfrg](https://tinyurl.com/y7vfsfrg)"
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE, comment=NA, fig.width=8, fig.height=3.4)
library(shiny)
```

Don't forget to load a library:
```{r}
library("lingtypology")
```

## 1. File management
### 1.1 Saving from R

There are two ways: base R and `readr` package. I mostly use the second way, since the base R functions haven't changed for decades, but for purpuses of simplicity will show you everything with base R functions.

```{r}
ejective_and_n_consonants
```

First we need to know `getwd()` and `setwd()` functions for setting a working directory.

```{r, eval = FALSE}
getwd()
setwd("/...") # use backslashes `\`, not slashes `/`!
```

So now we can save the dataframe:

```{r, eval = FALSE}
write.csv(ejective_and_n_consonants, "my_file_name.csv")
```

### 1.2 Loading data to R
We can read to R a dataset about Numeral Classifiers from [AUTOTYP database](https://github.com/autotyp/autotyp-data).
```{r}
new_df <- read.csv("https://raw.githubusercontent.com/autotyp/autotyp-data/master/data/Numeral_classifiers.csv")
new_df
```

It could be also a file on your computer, just provide a whole path to the file. Windows users need to change backslashes `\` to slashes `/`.

```{r, eval = FALSE}
new_df_2 <- read.csv("/home/agricolamz/my_file.csv")
```

### 1.3 Missing data
In R, missing values are represented by the symbol `NA` (not available), so ufter uploading all blanks that you have in your table will be changed to `NA`.

```{r}
is.na(c(5, 4, NA))
sum(is.na(c(5, 4, NA)))
```

## 2. Set pop-up boxes
Sometimes it is a good idea to add some additional information (e.g. language affiliation, references or even examples) to pop-up boxes that appear when points are clicked. So I create an extra vector of strings in the dataframe:
```{r}
df <- data.frame(language = c("Adyghe", "Kabardian", "Polish", "Russian", "Bulgarian"),
                 features = c("polysynthetic", "polysynthetic", "fusional", "fusional", "fusional"))
df$popup <- aff.lang(df$language)
```

The function `aff.lang()` creates a vector of genealogical affiliations that can be easily mapped:
```{r}
map.feature(languages = df$language, 
            features = df$features, 
            popup = df$popup)
```

Pop-up strings can contain HTML tags, so it is easy to insert a link, a couple of lines, a table or even a video and sound. Here is how pop-up boxes can demonstrate language examples:
```{r}
# change a df$popup vector
df$popup <- c("sɐ s-ɐ-k'ʷɐ <br> 1sg 1sg.abs-dyn-go <br> 'I go'",
              "sɐ s-o-k'ʷɐ <br> 1sg 1sg.abs-dyn-go <br> 'I go'",
              "id-ę <br> go-1sg.npst <br> 'I go'",
              "ya id-u <br> 1sg go-1sg.npst <br> 'I go'",
              "id-a <br> go-1sg.prs <br> 'I go'")

# create a map
map.feature(languages = df$language,
            features = df$features,
            popup = df$popup)
```

How to say _moon_ in Sign Languages? Here is an example:
```{r, echo=FALSE}
# Create a dataframe with links to video
sign_df <- data.frame(languages = c("American Sign Language", "Russian Sign Language", "French Sign Language"),
                      popup = c("https://media.spreadthesign.com/video/mp4/13/48600.mp4", "https://media.spreadthesign.com/video/mp4/12/17639.mp4", "https://media.spreadthesign.com/video/mp4/10/17638.mp4"))

# Change popup to an HTML code
sign_df$popup <- paste("<video width='200' height='150' controls> <source src='",
                       as.character(sign_df$popup),
                       "' type='video/mp4'></video>", sep = "")
# create a map
map.feature(languages = sign_df$languages, popup = sign_df$popup)
```

## 3. Set shapes
For some papers it is not possible to use colors for destinguishing features. In that cases it is posible to use `shape` argument:
```{r}
map.feature(languages = ejective_and_n_consonants$language,
            features = ejective_and_n_consonants$ejectives,
            shape = TRUE)
```

It is possible to provide a vector with corresponding shapes:
```{r}
map.feature(languages = ejective_and_n_consonants$language,
            features = ejective_and_n_consonants$ejectives,
            label = ejective_and_n_consonants$language,
            shape = c("–", "+"))
```

There is also possible to use numbers for pointing sumthing:
```{r, fig.height=4.2}
map.feature(languages = ejective_and_n_consonants$language,
            features = ejective_and_n_consonants$language,
            shape = 1:19) # 1:19 is a short way to say c(1, 2, 3, ..., 19)
```

## 4. Set coordinates
You can set your own coordinates using the arguments `latitude` and `longitude`. It is important to note, that `lingtypology` works only with decimal degrees (something like this: 0.1), not with degrees, minutes and seconds (something like this: 0° 06′ 0″). I will illustrate this with the dataset `circassian` built into the `lingtypology` package. This dataset comes from fieldwork collected during several expeditions in the period 2011-2016 and contains a list of Circassian villages:
```{r}
head(circassian)
```

In this dataframe you can find variables `latitude` and `longitude` that could be used:
```{r}
map.feature(languages = circassian$language,
            features = circassian$dialect,
            label = circassian$village,
            latitude = circassian$latitude,
            longitude = circassian$longitude,
            title = "Circassian dialects")
```

## 5. Set layouts
It is possible to use different tiles on the same map using the `tile` argument. For more tiles look [here](https://leaflet-extras.github.io/leaflet-providers/preview/index.html).
```{r}
map.feature(languages = circassian$language,
            features = circassian$dialect,
            label = circassian$village,
            latitude = circassian$latitude,
            longitude = circassian$longitude,
            tile = "OpenTopoMap")
```

It is possible to use different map tiles on the same map. Just add a vector with tiles.
```{r}
map.feature(languages = circassian$language,
            features = circassian$dialect,
            label = circassian$village,
            latitude = circassian$latitude,
            longitude = circassian$longitude,
            tile = c("OpenStreetMap.BlackAndWhite", "OpenTopoMap"))
```

## 6. Minicharts
The argument `minichart` allows you to add piecharts or barplots instead of standard point markers:
```{r}
map.feature(languages = ejective_and_n_consonants$language,
            minichart.data = ejective_and_n_consonants[, c("vowels", "consonants")],
            minichart = "bar")
map.feature(languages = ejective_and_n_consonants$language,
            minichart.data = ejective_and_n_consonants[, c("vowels", "consonants")],
            minichart = "pie")
map.feature(languages = ejective_and_n_consonants$language,
            minichart.data = ejective_and_n_consonants[, c("vowels", "consonants")],
            minichart = "pie",
            minichart.labels = TRUE)
```

## 7. Add a density contourplot to a map
Sometimes it is easier to look at a density contourplot. It can be created using `density.estimation` argument. There are two possibility for creation a density contourplot in `lingtypology`:

* `density.method = "fixed distance"`. First algorithm creates circle polygons with fixed radius around each point and then merge all polygons that are overlapped. It has only one parameter that should be estimated: radius of the circle (`density.width`).
* `density.method = "kernal density estimation"`. Second algorithm uses a kernal density estimation and has two parameters that should be estimated: latitude and longitude bandwidths (`density.width[1]` and (`density.width[2]`))


```{r}
map.feature(circassian$language,
            longitude = circassian$longitude,
            latitude = circassian$latitude,
            density.estimation = circassian$language,
            density.width = 0.15)
```

It is possible to remove points and display only the kernal density estimation plot, using the `density.points` argument:

```{r}
map.feature(circassian$language,
            longitude = circassian$longitude,
            latitude = circassian$latitude,
            density.estimation = circassian$language,
            density.width = 0.15,
            density.points = FALSE)
```

If you want to use kernal density estimation, you need to change method type and provide a vector of parameters that increase/decrease area:

```{r}
map.feature(circassian$language,
            features = circassian$language,
            longitude = circassian$longitude,
            latitude = circassian$latitude,
            density.estimation = "Circassian",
            density.method = "kernal density estimation",
            density.width = c(0.3, 0.3))

map.feature(circassian$language,
            features = circassian$language,
            longitude = circassian$longitude,
            latitude = circassian$latitude,
            density.estimation = "Circassian",
            density.method = "kernal density estimation",
            density.width = c(0.7, 0.7))

map.feature(circassian$language,
            features = circassian$language,
            longitude = circassian$longitude,
            latitude = circassian$latitude,
            density.estimation = "Circassian",
            density.method = "kernal density estimation",
            density.width = c(1.3, 0.9))
```

It is important to note, that this type of visualization have some shortcomings. The kernel density estimation is calculated without any adjustment, so longitude and latitude values used as a values in Cartesian coordinate system. To reduce consequences of that solution it is better to use a different coordinate projection. That allows not to treat Earth as a flat object.

## 8. Graticule

```{r}
map.feature("Adyghe",
            zoom.level = 12,
            graticule = 0.05) 
```

## 9. Minimap

```{r}
map.feature(c("Russian", "Adyghe"),
            minimap =  TRUE)
```

## 10. `mapview` package

The [`mapview` package](https://github.com/r-spatial/mapview) provides a possibility to create a multiple maps in a grid and even synchronise them. There are two functions for that: `latticeview()` and `sync()`. Facetisation is a really powerfull tool (look for `facet_grid()` and `facet_wrap()` functions from `ggplot2`). `lingtypology` doesn't provide a facetisation itself, but the `facet` argument of the `map.feature()` function create a list of maps based on this variable. The result of the work of this function then is changed: instead of creating a map in Viewer pane it will return a list that could be used in `latticeview()` and `sync()` functions from the `mapview` package.

```{r}
faceted <- map.feature(circassian$language,
                       latitude = circassian$latitude,
                       longitude = circassian$longitude,
                       features = circassian$dialect,
                       facet = circassian$language)
library(mapview)
sync(faceted, no.initial.sync = FALSE)
```

As you can see we provided a `circassian$language` to the `facet` argument, so it returned a list of two maps that stored in `faceted` variable.

It is also possible to combine any maps that were created, just store them in a variable, and combine them in `latticeview()` and `sync()` functions

```{r, eval = FALSE}
m1 <- map.feature(lang.aff("Tsezic"), label = lang.aff("Tsezic"))
m2 <- map.feature(lang.aff("Avar-Andi"), label = lang.aff("Avar-Andi"))
sync(m1, m2)
```

## 11. Create your own atlas with `rmarkdown`
This section is inspired by talk with [Niko Partanen](https://github.com/nikopartanen). It is possible to create an atlas website using `lingtypology` and [`rmarkdown`](https://github.com/rstudio/rmarkdown) packages. The function `atlas.database()` creates a folder in the working directory that contains an `rmarkdown` template for a web-site.

First, lets create a `dataframe` with some data.
```{r}
df <- wals.feature(c("1a", "20a"))
```

Second we can create a website using `atlas.database()` function:

* `languages` argument is a language list
* `features` argument is a data.frame with corresponding features
* `latitude` and `longitude` arguments are optional

```{r}
atlas.database(languages = df$language,
               features = df[,c(4:5)],
               latitude = df$latitude,
               longitude = df$longitude,
               atlas_name = "Some WALS features",
               author = "Author Name")
```

We can see that this function creates a subfolder with following files:
```{r}
list.files("./atlas_Some_WALS_features/")
```

The last step is to run a command:
```{r, eval=FALSE}
rmarkdown::render_site("./atlas_Some_WALS_features/")
```

Then the atlas website will be created (here is [a result](https://agricolamz.github.io/lingtypology_atlas_example/index.html)). If you want to change something in the website, just change some files:

* write information about atlas in index.Rmd file
* list citation information
* change any `.Rmd` file
* ...
* and on the end rerun the `rmarkdown::render_site("./atlas_Some_WALS_features/")` command.

```{r, include=FALSE}
unlink("./atlas_Some_WALS_features/", recursive = TRUE)
```
